---
title: "Long-term *Ipomopsis* Floral Volatiles"
author: "[John Powers](http://sites.uci.edu/powers/)"
date: "`r Sys.Date()`"
output:
  html_document:
    self_contained: no
    lib_dir: libs
    code_folding: hide
    toc: yes
    toc_float: TRUE
editor_options: 
  chunk_output_type: console
---
<style type="text/css">
.main-container { max-width: 1000px; margin-left: 0; margin-right: auto; }
img{ max-width:200%; height: auto; }
td, th { padding : 6px }
</style>

```{r setup, message=FALSE}
library(tidyverse)
library(pheatmap)
library(dendsort)
library(viridis)
library(RColorBrewer)
library(bouquet)
library(vegan)
library(ggvegan)
library(knitr)
knitr::opts_chunk$set(comment="", cache=T, warning = F, message = F, 
                      fig.path = "images/", dev="svglite", dev.args=list(fix_text_size=FALSE), fig.height=8, fig.width=8)
options(digits=4, knitr.kable.NA = "") # for kables
source("read_shimadzu.R")
explained <- function(cap) paste0(round(100*summary(cap)$constr.chi/summary(cap)$tot.chi, 0), "% explained")
```

```{r load_metadata}
metadata <- list.files("data/volatiles", pattern = "Ipomopsis long-term volatiles - 20", full.names=T) %>% 
  map_dfr(~read_tsv(.x, show_col_types = F) %>% 
            mutate(across(any_of(c("pump_id", "plant", "vial", "sample")), as.character)), .id="metafile") %>% 
  mutate(year = year(date))

# metadata %>% add_count(year, date, time, site, type, sample) %>% filter(n>1) %>% 
#   arrange(year, date, time, site, type, sample) %>% write_tsv("data/volatiles/metadupes.tsv", na="")
```

# Field metadata

```{r plot_metadata}
ggplot(metadata, aes(x=yday(date), fill=paste(site, time))) + 
  facet_wrap(vars(year), ncol=1) + geom_bar() +
  scale_fill_brewer(palette = "Paired") + theme_classic() + 
  labs(x="Day of year", y="Samples", fill="Site, time")

metadata %>% filter(type=="floral") %>% count(year, site, time) %>% 
  pivot_wider(names_from=c("site","time"), values_from="n") %>% kable(caption = "floral samples")
metadata %>% filter(type=="ambient") %>% count(year, site, time) %>% 
  pivot_wider(names_from=c("site","time"), values_from="n") %>% kable(caption = "ambient samples")
metadata %>% filter(type=="floral") %>% count(year, site, time, plant) %>% count(year, site, time) %>% 
  pivot_wider(names_from=c("site","time"), values_from="n") %>% kable(caption = "plants")

left_join(metadata %>% filter(type=="floral") %>% count(year, site, time, name="samples"),
          metadata %>% filter(type=="floral") %>% count(year, site, time, plant) %>% count(year, site, time, name="plants")) %>% 
  mutate(samples_per_plant = samples/plants, .keep="unused") %>% 
    pivot_wider(names_from=c("site","time"), values_from="samples_per_plant") %>% kable(caption = "samples per plant")
```

## Soil moisture

```{r soil_moisture}
metadata %>% filter(type=="floral") %>% group_by(year, site, time, date) %>% summarize(has_VWC = mean(!is.na(VWC))) %>% 
  ggplot(aes(y=fct_rev(paste(date, site, time)), color=paste(site, time), x=has_VWC)) + geom_point(size=3) +
  scale_color_brewer(palette = "Paired") + theme_minimal() + 
  labs(x="Fraction of floral samples with soil moisture recorded", y="Date, site, time", color="Site, time")

ggplot(metadata, aes(x=yday(date), color=paste(site, time), y=VWC)) + 
  facet_wrap(vars(year), ncol=1) + geom_point() +
  scale_color_brewer(palette = "Paired") + theme_classic() + 
  labs(x="Day of year", y="Soil moisture", color="Site, time")

ggplot(metadata, aes(x=site, fill=paste(site, time), y=VWC)) + 
  facet_wrap(vars(year), nrow=1) + geom_boxplot(outlier.size=0.5) +
  scale_fill_brewer(palette = "Paired") + theme_classic() + 
  labs(x="Site", y="Soil moisture", fill="Site, time")
```

## Extraction time

```{r time}
ggplot(metadata, aes(x=site, fill=paste(site, time), y=bag)) + 
  facet_grid(time~year, scales = "free_y") + geom_boxplot(outlier.size=0.5) +
  scale_fill_brewer(palette = "Paired") + theme_minimal() + 
  labs(x="Site", y="Bag time", fill="Site, time")

ggplot(metadata, aes(x=site, fill=paste(site, time), y=(pump-bag)/60)) + 
  facet_wrap(vars(year), nrow=1) + geom_boxplot(outlier.size=0.5) +
  scale_fill_brewer(palette = "Paired") + theme_minimal() + 
  labs(x="Site", y="Equilibration time (min)", fill="Site, time")

ggplot(metadata, aes(x=site, fill=paste(site, time), y=(end-pump)/60)) + 
  facet_wrap(vars(year), nrow=1) + geom_boxplot(outlier.size=0.5) +
  scale_fill_brewer(palette = "Paired") + theme_minimal() + 
  labs(x="Site", y="Pumping time (min)", fill="Site, time")
```

# Matching field metadata and filenames

```{r plot_matching}
filemeta <- read_csv("data/volatiles/Ipomopsis GCMS files - ipo_meta_split.csv") %>% 
  filter(type != "blank", #GC blank
         !verdict %in% c("skip-notrun", #tube was skipped by the autosampler
                         "leak-blank", #tube leaked, not desorbed but GC run anyway
                         "mismatch-leak", #tube leaked, tube time not aligned with file time
                         "skip-rename-mismatch-leak", #tube leaked after skip, tube time not aligned with file time
                         "filemoved" #file was moved on disk so creation time is wrong
         )) %>% drop_na(type, FileName) %>%  #no file match to desorb start time
  distinct(FileName, .keep_all = T) %>% #some files got duplicated from fuzzy matching times
  mutate(date = ymd(date), year=year(date), vial=as.character(vial))

#figured out which of these dupes has the real sample using PCA
filemeta %>% add_count(year, date, time, site, type, sample) %>% filter(n>1) %>% 
  arrange(year, date, time, site, type, sample) %>% write_tsv("data/volatiles/filedupes.tsv", na="")
filedupes <- read_tsv("data/volatiles/filedupes_annotated.tsv") %>% filter(!is.na(exclude)) %>% pull(FileName)
filemeta <- filemeta %>% filter(!FileName %in% filedupes)

bind_rows(files=filemeta, field=metadata, .id="source")%>%  
  ggplot(aes(y=paste(date, site, time), fill=source)) + 
  geom_bar(position=position_dodge(width=0.4)) + 
  scale_fill_brewer(palette = "Set2") + theme_classic() + 
  labs(y="Date, site, time", x="Samples", fill="Source")

full_join(count(filemeta, type, name="files"), count(metadata, type, name="field")) %>% kable()

full_join(count(filemeta, year, date, time, site, name="files"), 
          count(metadata, year, date, time, site, name="field")) %>% kable()
```

```{r matching}
#library(tidylog)
meta.full <- full_join(mutate(filemeta, files=1), 
                  mutate(metadata, field=1), by=c("year", "date", "time", "site", "type", "sample")) %>% 
  mutate(files=replace_na(files, 0), field=replace_na(field, 0), match = files==1 & field==1)

meta.full %>% filter(match==F) %>% arrange(date, time, site, field, files) %>% 
  select(date, time, site, type, field, files, sample, 
         rundate, samplename, vial.x, vial.y, split_notes, notes) %>% 
  write_tsv("data/volatiles/matching.tsv", na="")

meta.full %>% group_by(year, date, time, site) %>% 
  summarize(across(c(files, field, match), mean)) %>% 
  mutate(files=files-match, field=field-match) %>% 
  pivot_longer(c(files, field, match)) %>% 
  mutate(name=factor(name, levels=c("files", "match", "field"))) %>% 
  ggplot(aes(y=paste(date, site, time), x=value, fill=name)) + geom_col() +
  labs(x="",y="", fill="") + theme_minimal()
```

# Filtering compounds

```{r ipochems}
# load short names 
ipochems <- read_csv("data/volatiles/Ipo volatile compounds - chemsf_ipo.csv") %>% 
  select(name, shortname, standard, verdict) %>%  filter(verdict != "") %>% 
  mutate(standard = fct_recode(standard, "Methyl_salicylate"="Benzaldehyde"), # benzaldehyde regressions not reliable
         class = fct_recode(standard, Aliphatics="Hexenol", Benzenoids="Methyl_salicylate", Benzenoids="Indole", 
                            Sesquiterpenes="Caryophyllene", Monoterpenes="alpha_Pinene", Monoterpenes="Linalool")) %>%  
  left_join(read_csv("data/volatiles/regressions_181921_filtered_slopes.csv") %>% 
              pivot_wider(id_cols=standard, names_from="year", names_prefix="area_per_ng", values_from=area_per_ng)) 
class_pal <- set_names(c("#BC0060","#027B8C","#E56E00","#86D400"), levels(ipochems$class))

#shorten chemical names and merge compounds with multiple names
shortnames <- ipochems %>% select(name, shortname) %>% filter(shortname!="") %>% deframe()
#shortnames[shortnames %in% shortnames[duplicated(shortnames)]]

greekify <- function(names) {
  names %>% 
    str_replace("^a-","\U03B1-") %>% str_replace("-a-","-\U03B1-") %>% 
    str_replace("^b-","\U03B2-") %>% str_replace("-b-","-\U03B2-") %>% 
    str_replace("^g-","\U03B3-") %>% str_replace("-g-","-\U03B3-")
}
```


```{r quant}
quant.full <- list.files("data/volatiles/quant/", full.names = T) %>% 
  map_dfr(read.shimadzu.quant) %>%
  mutate(Name = trimws(Name), Area=replace_na(Area, 0)) %>% 
  distinct(Filename, Name, .keep_all=T)

need_meta <- setdiff(quant.full$Filename, meta.full$FileName)#TODO no metadata, consider filling some in

#only keep file-metadata matches
files.match <- intersect(meta.full$FileName, quant.full$Filename)
meta.all   <- meta.full %>% filter(FileName %in% files.match)
quant.all <- quant.full %>% filter(Filename %in% files.match) %>%
  select(Filename, Name, Area) %>%
  mutate(Name = recode(Name, !!!shortnames)) %>%
  pivot_wider(names_from = "Name", values_from="Area") %>% 
  arrange(match(Filename, meta.all$FileName)) %>% 
  write_tsv("data/volatiles/quant_all.tsv") %>% 
  as.data.frame() %>% column_to_rownames("Filename")
```

```{r checkdupes, eval=FALSE}
filedupes <- read_tsv("data/volatiles/filedupes_annotated.tsv") %>% 
  mutate(fullsample = paste(year, date, time, site, type, sample), exclude=!is.na(exclude)) %>% 
  rename(Filename=FileName) %>% left_join(quant.full)
filedupes.wide <- filedupes %>% 
  select(Filename, Name, Area) %>%
  mutate(Name = recode(Name, !!!shortnames)) %>%
  filter(Name !="1,6,10-Dodecatrien-3-ol, 3,7,11-trimethyl-") %>% 
  pivot_wider(names_from = "Name", values_from="Area") %>% 
  as.data.frame() %>% column_to_rownames("Filename")
filedupes.meta <- filedupes %>% distinct(Filename, fullsample, .keep_all = T)

filedupes.pca.df <- fortify(rda(sqrt(filedupes.wide)))
ggplot(filter(filedupes.pca.df, score=="sites"), aes(x=PC1, y=PC2, color=factor(filedupes.meta$year))) +
  geom_vline(xintercept = 0) + geom_hline(yintercept = 0)+
  geom_text(data=filter(filedupes.pca.df, score=="species"), aes(label=greekify(label)), color="grey40") +
  geom_path(aes(group=filedupes.meta$fullsample)) + geom_point() + 
  geom_text(aes(label=if_else(filedupes.meta$exclude, "", str_remove(filedupes.meta$Filename,".qgd")))) + 
  guides(color="none") + theme_classic()
```


```{r qual}
qual.full <- list.files("data/volatiles/qual/", full.names = T) %>% 
  map_dfr(read.shimadzu)

#only keep file-metadata matches
files.match <- intersect(meta.full$FileName, qual.full$Filename)
meta.all   <- meta.full %>% filter(FileName %in% files.match)
qual.data <-  qual.full %>% filter(Filename %in% files.match) %>% 
  filter(Name != "") %>% #drop peaks without identifications
  mutate(Name = droplevels(recode(Name, !!!shortnames)))#bouquet doesn't like empty levels
qual.all <- qual.data %>% select(Filename, Name, Area) %>%
  pivot_wider(names_from = "Name", values_from="Area", values_fn = sum) %>% 
  arrange(match(Filename, meta.all$FileName)) %>% 
  as.data.frame() %>% column_to_rownames("Filename")
```

```{r bouquetprep}
longdata <- qual.data %>% load_longdata(sample="Filename", RT="Ret.Time", 
                          name="Name", area="Area", match = "SI", maxmatch=100) 
metadata.df <- meta.all %>% 
  as.data.frame() %>% #tibbles break bouquet's add_count_freqs
  select(-sample) %>% #name conflict with bouquet
  load_metadata(date="date", sample="FileName", type="type")
rownames(metadata.df) <- metadata.df$sample

qual.all <- make_sampletable(longdata, metadata.df)
chems <- make_chemtable(longdata, metadata.df)
```


## Correlations among unfiltered compounds 

```{r corr_all, fig.height=25, fig.width=12, dev="png", dev.args=list()}
meta.all.noleaf <- meta.all %>% filter(type != "leaf")
cut_threshold <- 5e8
qual.all.cut <- qual.all[meta.all.noleaf$FileName, colSums(qual.all) > cut_threshold | #big compounds
                           colnames(qual.all) %in% colnames(quant.all)] #or in unfiltered quant list

contams <- c("Nonanoic acid", "Decanal", "Caprolactam")
inquant <- data.frame(inquant=if_else(colnames(qual.all.cut) %in% setdiff(colnames(quant.all), contams),"Y","N"),
                      row.names=colnames(qual.all.cut))
pheatmap(cor(sqrt(qual.all.cut)), scale="none", 
         clustering_method = "mcquitty", cutree_rows = 20,
         clustering_distance_rows="correlation", clustering_distance_cols="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         annotation_col = inquant, annotation_row = inquant,
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01), 
         labels_row=greekify(colnames(qual.all.cut)), show_colnames = F)
```

## Check ambients vs. samples

```{r check_ambi}
cap.ambi <- capscale(sqrt(qual.all.cut) ~ type, data=meta.all.noleaf)
anova(cap.ambi)
explained(cap.ambi)

cap.ambi.df <- fortify(cap.ambi) %>% 
  left_join(mutate(meta.all.noleaf, score="sites"), by=c(label="FileName", "score")) %>% 
  left_join(rownames_to_column(inquant, "label"))

zoom <- 40
cap_threshold <- 3
ggplot(filter(cap.ambi.df, score=="sites"), aes(x=CAP1, y=MDS1, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top") + 
  geom_point(aes(color=type, shape=type)) + scale_shape_manual(values=c(19,1))+
  scale_color_manual(values=c("darkblue","magenta")) +
  geom_text(data=filter(cap.ambi.df, score=="species", abs(CAP1)>cap_threshold), aes(x=CAP1*zoom))

#filter(cap.ambi.df, score=="species", label%in%colnames(quant.all)) %>% arrange(CAP1) %>% View()
ggplot(filter(cap.ambi.df, score=="species"),  aes(x=CAP1, y=MDS1, color=inquant)) +
  labs(color= "In quant integrations?")+
  theme_classic() + theme(legend.position = "top") +
  geom_vline(xintercept=0)+ geom_hline(yintercept=0)+
  geom_smooth() + geom_point() 

compounds.cap.angles <- cap.ambi.df %>% filter(score=="species") %>% 
  mutate(angle = atan2((CAP1*zoom), MDS1)*180/pi)

ggplot(compounds.cap.angles, aes(x=angle, fill=inquant)) + facet_wrap(vars(inquant), ncol=1) +
  geom_histogram(binwidth=5) + scale_x_continuous(breaks=seq(-180, 180, by=30))+
  labs(fill= "In quant integrations?")

# cap.candidates <- compounds.cap.angles %>% 
#   filter(abs(angle+90)<5, inquant=="N") %>% select(name=label, CAP1, angle) %>% arrange(CAP1)
```


## Filtering criteria

```{r bouquet}
chemsf <- chems %>%
  filter_RT(2, 17) %>% 
  filter_match(0.8) %>% 
  filter_freq(0.1) %>% 
  filter_contaminant(cont.list = c("Caprolactam")) %>%
  filter_area(min_maximum = 1e5) %>% 
  filter_ambient_ratio(qual.all, metadata.df, ratio = 4) %>%
  combine_filters() 

chemsf$filter_final <- with(chemsf, filter_freq.floral=="OK" &
                              filter_ambient_ratio=="OK") #only use these for final filtering

bouquet::plot_filters(chemsf, option="rarity", yrange=3.5)
bouquet::plot_filters(chemsf, option="ambient", yrange=3.5)
bouquet::plot_filters(chemsf, option="prop")
```

## Compare quant and filtered compounds

```{r quant_qual}
qual_chems <- chemsf$name[chemsf$filter_final]

#TODO check on new compounds to investigate
(qual_only <- setdiff(qual_chems, names(quant.all)))
#chemsf %>% filter(filter_final) %>% mutate(check = name %in% qual_only, .before=1) %>% View()

#TODO see if any other quant peak should be included
(quant_only <- setdiff(names(quant.all), qual_chems))
#chemsf %>% filter(filter_final) %>% mutate(check = name %in% iponew, .before=1) %>% View()

# compounds in both quant and qual lists
(quant_qual <- intersect(names(quant.all), qual_chems))

toolow <- rownames(quant.all)[rowSums(quant.all[,quant_qual]) < 2500] #exclude 2 samples with very low integrations

qual <- prune_sampletable(qual.all, chemsf, metadata.df)[, quant_qual]
qual <- qual[!rownames(qual) %in% toolow,]
quant <- quant.all[rownames(qual), quant_qual]
quant.long <- quant %>% rownames_to_column("sample") %>% pivot_longer(colnames(quant))
ipochemsf <- ipochems[match(colnames(qual), ipochems$shortname),]
meta <- metadata.df[rownames(qual),] %>% 
  mutate(across(c(year, site, time), factor))

quant %>% rownames_to_column("FileName") %>% write_tsv("data/volatiles/filtered_quant.tsv")
meta %>% write_tsv("data/volatiles/filtered_meta.tsv")
save(quant, meta, file = "data/volatiles/filtered_quantmeta.rda")
```

## Retention time of qual integrations

```{r retention}
ggplot(longdata %>% filter(name %in% qual_chems), 
       aes(x=RT, y=fct_reorder(name, RT), color=factor(name, levels=sample(levels(name))))) + 
  geom_jitter(width = 0, height=0.5) + guides(color="none") + ylab("")
#TODO make sure quant has correct RT, investigate second peaks
#TODO cosmene/carveol/alloocimene puzzle
#TODO (E)-hex-2-enal vs. no isomer specified
```

# Descriptions of filtered volatiles

## Sample size

```{r sampsize}
meta %>% count(year, site, time) %>% 
  pivot_wider(names_from=c("site","time"), values_from="n") %>% kable(caption = "floral samples")
meta %>% count(year, site, time, plant) %>% count(year, site, time) %>% 
  pivot_wider(names_from=c("site","time"), values_from="n") %>% kable(caption = "plants")
```


## Correlations

```{r corr, fig.width=10, fig.height=8}
corr.annot <- ipochemsf %>% select(shortname, Class=class) %>% column_to_rownames("shortname")
corr.vols <- cor(quant, method="pearson")
pheatmap(corr.vols, 
         scale="none", clustering_method = "mcquitty",
         color=colorRampPalette(c("blue","white","red"))(200), breaks=seq(-1,1,by=0.01),
         annotation_col=corr.annot, annotation_row=corr.annot, 
         annotation_colors = list(Class=class_pal),
         labels_row=greekify(rownames(corr.vols)), labels_col=greekify(colnames(corr.vols)))
```

## Heatmap

```{r quant_heatmap, dev="png", dev.args=list(),fig.width=11, fig.height=8}
load("data/volatiles/filtered_quantmeta.rda")

pal <- with(meta, list(
  year = set_names(brewer.pal(nlevels(year), "Set2"), levels(year)),
  site = set_names(brewer.pal(nlevels(site), "Set1"), levels(site)),
  time = set_names(c("lightblue","plum1"), levels(time))))
times <- c(D="day", N="night")
dots <- c(D=1, N=19)

pheatmap(as.matrix(t(quant))^(1/4), 
         cluster_cols=T, show_colnames=F,
         clustering_method="mcquitty", clustering_distance_rows="correlation",
         clustering_callback = function(hc, ...){dendsort(hc, type="average")},
         annotation_col = meta %>% select(year, site, time) %>% mutate(across(everything(), factor)),
         annotation_colors =pal,
         scale="none", color=mako(512),main="Filtered quant volatiles")
```

## Boxplots

```{r boxplots, fig.width=14, fig.height=10}
quant.long %>% left_join(meta) %>% mutate(name=fct_reorder(greekify(name), -value)) %>% 
  ggplot(aes(x=year, y=value, fill=time, color=site)) + facet_wrap(vars(name), scales="free_y") + 
  geom_boxplot(outlier.size= 0.1) + scale_y_sqrt() +
  scale_color_manual("Site", values=pal$site) + scale_fill_manual("Time", values=c(D="white",N="black")) +
  theme_bw() + theme(axis.title=element_blank(), legend.position = "top")
```

## NMDS

```{r quant_nmds}
nmds <- metaMDS(decostand(quant, "hellinger"), autotransform = F, trymax=1, trace=F)

nmds.df <- fortify(nmds) %>% left_join(mutate(meta, score="sites"), by=c(label="sample", "score"))
nmds.plot <- ggplot(filter(nmds.df, score=="sites"), aes(x=NMDS1, y=NMDS2, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top")

nmds.plot + geom_point(aes(color=time)) +
  geom_text(data=filter(nmds.df, score=="species")) +
  scale_color_manual("Time", values=pal$time, label=times)
  
nmds.plot + geom_point(aes(color=site, shape=time)) +
  scale_color_manual("Site", values=pal$site) + 
  scale_shape_manual("Time", values=dots, label=times)

nmds.plot + geom_point(aes(color=year, shape=time)) +
  geom_text(data=filter(nmds.df, score=="species")) +
  scale_color_manual("Year", values=pal$year) + 
  scale_shape_manual("Time", values=dots, label=times)
```

# Test group differences

## CAP with time

```{r cap_time}
cap.t <- capscale(sqrt(quant) ~ time, data=meta)
explained(cap.t)

cap.t.df <- fortify(cap.t) %>% left_join(mutate(meta, score="sites"), by=c(label="sample", "score"))

ggplot(filter(cap.t.df, score=="sites"), aes(x=CAP1, y=MDS1, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top") + 
  geom_vline(xintercept = 0)+
  geom_point(aes(color=year, shape=time)) +
  geom_text(data=filter(cap.t.df, score=="species")) +
  scale_color_manual("Year", values=pal$year) + 
  scale_shape_manual("Time", values=dots, label=times)

ggplot(filter(cap.t.df, score=="sites"), aes(y=-CAP1, x=pump, color=year)) + 
  geom_point(shape=1) + geom_smooth(aes(group=paste(time, year)), se=F, span=1.4) +
  scale_color_manual(values=pal$year) + theme_minimal() +
  scale_x_time(breaks=as.POSIXct(seq(0,24, by=2)*60*60), labels=hour) + 
  labs(x="Hour of sampling", y="CAP separating day and night", color="Year")
```

## CAP with time, year, site

```{r cap_tys}
cap.tys <- capscale(sqrt(quant) ~ time + year + site + year:site + time:site, data=meta)
anova(cap.tys, by="term")
explained(cap.tys)

cap.tys.df <- fortify(cap.tys) %>% left_join(mutate(meta, score="sites"), by=c(label="sample", "score"))

ggplot(filter(cap.tys.df, score=="sites"), aes(x=CAP1, y=CAP2, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top") + 
  geom_point(aes(color=year, shape=time)) +
  geom_text(data=filter(cap.tys.df, score=="species")) +
  scale_color_manual("Year", values=pal$year) + 
  scale_shape_manual("Time", values=dots, label=times)
```

## CAP with time, year, site, soil moisture

```{r cap_vwc}
meta.withvwc <- meta %>% drop_na(VWC) %>% 
  group_by(date, site, time) %>% #see plot above for which dates have soil moisture data
  mutate(relVWC = VWC - mean(VWC)) %>% ungroup() #VWC centered to mean for each sampling date
quant.withvwc <- quant[meta.withvwc$sample,]

cap.tysv <- capscale(sqrt(quant.withvwc) ~ time + year + site + relVWC, data=meta.withvwc)
anova(cap.tysv, by="term")
explained(cap.tysv)
```

## CAP with year, within time

```{r cap_year}
meta.D <- filter(meta, time=="D")
quant.D <- quant[rownames(meta.D),]
meta.N <- filter(meta, time=="N")
quant.N <- quant[rownames(meta.N),]

cap.year.D.df <- capscale(sqrt(quant.D) ~ year, data=meta.D) %>% 
  fortify() %>% left_join(mutate(meta, score="sites"), by=c(label="sample", "score"))
ggplot(filter(cap.year.D.df, score=="sites"), aes(x=CAP1, y=CAP2, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top") + 
  geom_point(aes(color=year, shape=time)) +
  geom_text(data=filter(cap.year.D.df, score=="species")) +
  scale_color_manual("Year", values=pal$year) + 
  scale_shape_manual("Time", values=dots, label=times)

cap.year.N.df <- capscale(sqrt(quant.N) ~ year, data=meta.N) %>% 
  fortify() %>% left_join(mutate(meta, score="sites"), by=c(label="sample", "score"))
ggplot(filter(cap.year.N.df, score=="sites"), aes(x=CAP1, y=CAP2, label=greekify(label))) +
  theme_classic() + theme(legend.position = "top") + 
  geom_point(aes(color=year, shape=time)) +
  geom_text(data=filter(cap.year.N.df, score=="species")) +
  scale_color_manual("Year", values=pal$year) + 
  scale_shape_manual("Time", values=dots, label=times)
```
